---
title: "Arm or Legs? Outfield Range and Throwing Value"
author: "Leo Foust"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    theme: readable
    mathjax: null
---

```{r setup, include=FALSE}
# reading in the saved results and setting up the report
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE,fig.width=9,fig.height=5)
library(tidyverse)
read_table <- function(path) read_csv(file.path('data/processed',path),show_col_types=FALSE)
models <- read_table('models/model_comparison.csv')
components <- read_table('jump_components/model_comparison.csv')
attempts <- read_table('runner_attempts/model_comparison.csv')
odds <- read_table('runner_attempts/arm_odds_ratio.csv')
samples <- read_table('samples/sample_counts.csv')
disp <- read_table('models/component_dispersion.csv')
unc <- read_table('final/dispersion_uncertainty.csv')
sensitivity <- read_table('sensitivity/model_scores.csv')
include_plot <- function(path) knitr::include_graphics(normalizePath(path),rel_path=FALSE)
num <- function(x) formatC(x,format='f',digits=2)
```

<style>
table {display:block; overflow-x:auto; font-size:13px;}
th {white-space:normal;}
img {max-width:100%; height:auto;}
body {line-height:1.65;}
</style>

# Main takeaways

In this saved 2025 sample, **range runs vary more between players than throwing
runs**, overall and within primary-position groups. Jump and Burst carry more
information about range value than sprint speed in the fitted models. OF arm
strength adds some information about throwing value. Evidence that it materially
improves prediction of runner attempts is weak.

The question here is about what happened in this saved season. When I say one
component “matters more,” I am looking at how much it separates the players in
these measurements. That is different from proving how much a training program
would help or predicting next season. The analysis also keeps track of how range
and arm values vary together, rather than turning them into percentages of each
player's total.

## Important audit correction

The original Arm Strength export lacked a year column and did **not** match the
official 2025 leaderboard. Early versions of the throwing analyses used it.
Those throwing results are superseded. The current pipeline uses
`arm_strength_2025_verified.csv`, extracted from records explicitly tagged 2025.
The original file remains unchanged for provenance. This report and current CSV
outputs are the authoritative results; historical discussion summaries are not.

# Where the data came from

The data come from five Statcast exports and three FanGraphs files for positional
innings. The code joins players by ID because names can be written differently.
It converts baseball innings into outs before adding them and leaves missing
measurements as missing. If a player is absent from a positional innings file,
that is treated as zero innings at that position, based on how the files were
collected. Records with no outfield innings stay in the data checks but are not
used in the analysis.

A player needs at least 60% of his outfield innings at one position to receive
that position label. Otherwise, he is in the multi-position group. Each question
uses the available players, but competing models within that question always use
the same group.

```{r sampletable}
# showing the number of players used for each question
knitr::kable(samples |> filter(sample %in% c('range_speed_and_components','arm_value_of_strength','deterrence_of_strength','range_vs_arm','range_vs_arm_by_position')) |> select(sample,included),col.names=c('Analysis availability','Players'))
```

Run `source("run_all.R")` from the RStudio project or `Rscript --vanilla run_all.R`
from its directory. The pipeline checks SHA-256 hashes before and after, runs
each script in a separate environment, saves generated outputs and records R and
package versions. It installs nothing and needs no network access for reruns.
Recorded versions describe the successful environment; they are not a dependency lockfile.

## Source verification

On 2026-10-01, the official 2025 qualified OAA roster matched all 110 saved players,
with qualification flags and identical range runs. The qualified Arm Value roster
matched all 118 players and opportunity counts. Arm run values had small revisions
(maximum absolute difference 0.0713 runs); the analysis preserves the original
snapshot. Sprint Speed's 579 players, displayed speeds and competitive-run counts
matched the official 2025 page. Arm Strength was replaced as described above.

Jump's counts satisfy its published rule under the stated full-season assumption.
FanGraphs year provenance rests on the original collection context and filenames;
the files have no season column. Matching modern snapshots cannot reconstruct
all historical filters. The numerical Arm Value qualification formula remains
undocumented in the inspected public interface; matching its qualified roster
provides membership evidence without inventing that formula.

# Range: speed or movement?

Every range model includes outfield innings and position. Innings help account
for playing time, although they do not tell us how difficult a player's chances
were. To check a model, the code leaves one player out, fits it on everyone else,
and predicts that player. Then it repeats for the others. RMSE gives more weight
to large misses; MAE is the average size of a miss. Both scores are in runs, and
lower is better.

```{r range}
# comparing prediction errors for the range models
knitr::kable(models |> filter(family=='range') |> select(model,players,loocv_rmse,loocv_mae),digits=3)
knitr::kable(components |> select(model,players,loocv_rmse,loocv_mae),digits=3)
include_plot('figures/09_jump_component_comparison.png')
```

Jump helped explain range more than sprint speed alone did. Burst had the best
score among the individual movement components, but its small advantage over
Jump does not prove it is always better. Reaction and Route had a correlation of
about −0.85, so their separate effects are hard to untangle when both are in the
model. The models use either overall Jump or its components, not both together.

These comparisons were explored after looking at the data. Testing several
models can make the best-looking result seem stronger than it really is, even
with cross-validation. The analysis does not include an uncertainty interval for
the difference between model errors or a test on another season.

# Throwing value

These models use the corrected 2025 outfield arm-strength data. They include
advancement opportunities and position before adding strength. The comparison
using overall arm strength keeps the same players so that changing the sample
does not also change the comparison.

```{r arm}
# comparing prediction errors for the arm models
knitr::kable(models |> filter(family=='arm') |> select(model,players,loocv_rmse,loocv_mae),digits=3)
include_plot('figures/03_arm_value.png')
```

Savant's throwing-value calculation already uses throwing arm among its inputs.
That means the two measurements are connected by design. A model coefficient
shows a relationship after accounting for the other included variables; it does
not tell us how many runs a player would gain by training to throw harder.

# Runner attempts

The runner-attempt models use the number of attempts and non-attempts. This keeps
track of how many opportunities each player's rate is based on. Both models
include position, and the second adds outfield arm strength. This type of model
is called quasibinomial. It allows the variation in the counts to differ from a
standard binomial model. The arm-strength effect is assumed to be the same across
position groups.

```{r attempts}
# showing the runner-attempt model results
knitr::kable(attempts |> mutate(across(c(player_rate_rmse,player_rate_mae),~100*.x)),digits=5,
 caption='Player-rate RMSE and MAE are percentage points; event scores are unitless.')
knitr::kable(odds,digits=3)
include_plot('figures/11_runner_attempts.png')
```

The odds ratio describes the estimated change in odds for 5 more mph of arm
strength. It is not a percentage-point change in attempt rate, and its interval
depends on this model's assumptions. Runner speed, ball location, bases and outs,
parks, and other play details are not included. The score differences are too
small to conclude that arm strength meaningfully improved prediction. Giving
each player equal weight and giving each opportunity equal weight answer slightly
different questions.

# Range versus arm on the same players

```{r dispersion}
# comparing the spread of range and arm runs
knitr::kable(disp |> filter(scale=='season_totals'),digits=3)
include_plot('figures/05_range_vs_arm.png')
include_plot('figures/12_dispersion_uncertainty.png')
```

Total measured runs equal range runs plus arm runs. The spread of that total
also depends on whether the two components tend to move together. In statistical
terms: Var(T) = Var(R) + Var(A) + 2 Cov(R,A). That is why the table includes
covariance. A larger range spread does not mean every player gets more value
from range or that range is worth a fixed multiple of throwing. Position groups
use each player's full-season outfield values, including work at other positions.

To estimate uncertainty in the spread, the code repeatedly samples players with
replacement 2,000 times within each group. It keeps a player's range and arm
values together. This is called a paired bootstrap. The intervals describe
uncertainty within this selected sample. They do not account for every measurement
problem, missing player, qualification rule, or change from season to season.

# Checking whether the findings held up

```{r sensitivity}
# showing whether the results hold up under different sample rules
knitr::kable(read_table('sensitivity/component_dispersion.csv'),digits=3)
knitr::kable(read_table('sensitivity/influence_summary.csv'),digits=5)
include_plot('figures/07_model_residuals.png')
include_plot('figures/08_model_qq.png')
```

I checked whether the position rule changed the results by using 50%, 60%, and
70% cutoffs. Players stay in the overall models while their labels may change.
The position-specific comparisons require the chosen share of innings.

The code also flags players who have a large effect on a fitted model, using
Cook's distance greater than 4/n, where n is the sample size. It then repeats the
analysis without each flagged player and without all of them together. Competing
models still use the same players in each check. These are ways to test the
findings, not reasons to delete players just because the results look better
without them. The checks change one choice at a time.

Range still had a larger spread after removing each of the 107 players one at
a time. Other checks compared runs per 1,000 innings and the values left after
accounting for innings with a linear model. Neither approach fully accounts for
different defensive opportunities. Some unusually large model errors remained.
The saved linear-model results include HC3 standard errors, which allow for
unequal error variation. No player was automatically removed for being influential.

# What this means for baseball

## Completing the original position question

The original question also asked whether the balance changes between LF, CF,
and RF. The extension compares the spread of range and arm runs per 450 innings
and directly estimates differences between positions. It uses the existing
60% position rule without a new innings minimum as its primary sample.

```{r position_extension}
# comparing the range and arm balance at each position
balance <- read_table('original_questions/position_balance.csv')
contrasts <- read_table('original_questions/position_contrasts.csv')
knitr::kable(balance |> filter(scenario %in% c('primary','minimum_450')) |>
 select(scenario,position,players,range_sd,arm_sd,sd_ratio),digits=3)
knitr::kable(contrasts |> filter(scenario=='primary') |>
 select(contrast,balance_ratio,lower_95,upper_95,lower_family,upper_family),digits=3)
include_plot('figures/13_position_balance.png')
```

The ratio is range standard deviation divided by arm standard deviation. Values
above 1 mean range varies more; they are not value multipliers. In the primary
sample, the ratio is 3.82 in CF, 2.18 in LF, and 1.95 in RF. The CF/RF contrast
is 1.96, with a 95% interval of 1.03–3.49. A wider interval that accounts for the
three planned comparisons is 0.90–3.96, which includes no difference.

With a 450-inning minimum, the CF/RF contrast shrinks to 1.14. This makes the
position hierarchy uncertain. It does not mean positions are equivalent.
Range has the larger point-estimate spread in every scenario; the 900-inning LF
sample is small enough that its interval includes equal spreads.

The extension uses 5,000 paired-player bootstrap draws within each position.
The three contrast intervals use a 98.33% percentile interval as an approximate
Bonferroni adjustment. Uncertainty remains conditional on the selected players;
the data do not isolate the effect of changing a player's position.

## Position interactions and arm-strength curvature

An interaction allows a skill's relationship with the outcome to differ by
position. These comparisons use the same LF/CF/RF players for every model within
each question: 77 for range and 98 for arm value and runner attempts.

```{r interaction_extension}
# showing whether separate relationships by position improve predictions
knitr::kable(read_table('original_questions/interaction_scores.csv') |>
 filter(scenario=='primary',group=='Overall') |>
 mutate(rmse=if_else(family=='attempts',100*rmse,rmse),
        units=if_else(family=='attempts','percentage points','runs')) |>
 select(family,model,players,rmse,units),digits=3)
knitr::kable(read_table('original_questions/arm_curvature_scores.csv') |>
 filter(group=='Overall') |> select(model,players,rmse,mae),digits=3)
```

Position improved the Jump model modestly, but separate position slopes did not
improve range or arm-value RMSE over shared slopes. Separate slopes made a small
improvement for runner attempts. The RF strength–attempt slope was negative,
but its interval adjusted for checking three position slopes included zero.
This is tentative evidence rather than a strong deterrence result.

The quadratic arm model reduced RMSE from 1.912 to 1.907 runs. That is too small
an improvement to establish a useful velocity threshold. A quadratic check does
not test every possible nonlinear relationship.

## What this means for choosing a player

At equal exposure, a one-run range advantage offsets a one-run arm disadvantage.
It is the size of both differences that matters. Among CFs with at least 450
innings, Victor Scott II's range advantage over Brenton Doyle was 3.97 runs per
450 innings and his arm disadvantage was 3.36, leaving him 0.61 ahead overall.
Michael Harris II's range advantage over Doyle was only 0.19, against an arm
disadvantage of 2.42, leaving him 2.23 behind. These are observed 2025 rates,
not predictions of who would be the better future acquisition.

The hold, advance, and thrown-out components reconcile to arm runs for all 118
players. The data can credit holds without a recorded out, but do not establish
that arm strength caused those holds. For example, Michael Harris II had about
+5.40 hold runs and −6.39 advance runs, with zero recorded outs in the tracked
advancement data, for about −0.99 total arm runs.

A question-by-question explanation is saved in
`reports/original_questions_answered.md`, and all extension outputs are in
`data/processed/original_questions/`. The extension was specified after the
initial findings and remains exploratory.

## Overall interpretation

The baseball takeaway is to look at how an outfielder moves toward the ball,
not just how fast he can run. Throwing needs its own outcome because OAA measures
range. This project does not establish that a right fielder's arm has a larger
causal value than a left fielder's, tell a player how to train, or provide enough
evidence for an acquisition decision.

The project can be rerun from the saved 2025 data. A next step would be to add
another season and more detail about the plays each player faced. That would
help test whether the relationships hold up outside this sample.

# Sources

- [Baseball Savant OAA](https://baseballsavant.mlb.com/leaderboard/outs_above_average)
- [Outfielder Jump](https://baseballsavant.mlb.com/jump)
- [Arm Strength](https://baseballsavant.mlb.com/leaderboard/arm-strength)
- [Arm Value](https://baseballsavant.mlb.com/leaderboard/baserunning?type=Fld)
- [Sprint Speed](https://baseballsavant.mlb.com/leaderboard/sprint_speed)
- [FanGraphs fielding leaderboards](https://www.fangraphs.com/leaders/major-league?stats=fld&season=2025&season1=2025)

Exact verification URLs and extracted public source records are saved in
`docs/verification`. Original source hashes are in `docs/source_manifest.csv`.

# Validation

```{r checks}
# showing the saved validation checks
knitr::kable(read_table('final/validation.csv'))
```
